Switch equivalence and weak prime domains for fusions

### 2023-12-11 fusione.tex
indep is global + varie

\usepackage[all]{xy}
\SelectTips{cm}{}
% derivations and labels
\usepackage{proof}
\newcommand{\lab}[1]{\ensuremath{\mathsf{{#1}}}}
\newcommand{\slab}[1]{\ensuremath{\scriptstyle{\mathsf{{#1}}}}}

@@

%identità
\newcommand{\id}[1]{\mathsf{id}_{#1}}
%codominio
\newcommand{\cod}[1]{\mathsf{cod}({#1})}
	%Variabili categorie
\def\A{\textbf {\textup{A}}}
%scheletro

@@

\title{Switch equivalence and weak prime domains for fusions}
\author{Paolo Baldan \and Davide Castelnovo  \and  Andrea Corradini  \and Fabio Gadducci}

@@

This first section is devoted to recall the definition and the basic theory of \emph{$\mathcal{M}$-adhesive categories} \cite{azzi2019essence,ehrig2006weak,lack2005adhesive}.

\begin{notation} 
We stipulate here some notational conventions which will be used throughout this paper.

Given a category $\X$ we will not distinguish notationally between $\X$ and its class of objects: so that ``$X\in \X$'' means that $X$ belongs to the class of objects of $\X$.

If $1$ is a terminal object in a category $\X$,  the unique arrow $X\to 1$ from another object $X$ will be denoted by $!_X$. Similarly, if $0$ is initial in $\X$ then $?_X$ will denote the unique arrow $0\to X$. When $\X$ is $\Set$ and $1$ is a singleton, $\delta_x$ will denote the arrow $1\to X$ with value $x\in X$.

Finally, we will use the following notation for some special classes of arrows of a category $\X$:
\begin{itemize}
	\item $\mathcal{A}(\X)$ will denote the class of all arrows of $\X$;
	\item $\mathcal{M}(\X)$ will denote the class of all monos of $\X$;
	\item $\mathcal{R}(\X)$ will denote the class of all regular monos of $\X$.
\end{itemize}
\end{notation}

\subsection{The Van Kampen condition}
The key property that $\mathcal{M}$-adhesive categories enjoy is given by  the so-called \emph{Van Kampen condition} \cite{brown1997van,johnstone2007quasitoposes,lack2005adhesive}. We will recall it and examine some of its consequences.

\begin{definition} \index{Van Kampen square} \index{pushout!- square} \index{pushout!Van Kampen -|see{Van Kampen square}}Let $\X$ be a category  and consider the two diagrams below
	\[\xymatrix@C=10pt@R=10pt{&&&&&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar[dl]_{m'} && B' \ar[dd]^{b} \ar[dl]_{n'} \\ A \ar[dd]_{m}\ar[rr]^{f}&&B\ar[dd]^{n}&&C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&&&&&A\ar[rr]|\hole^(.65){f} \ar[dl]^{m} && B \ar[dl]^{n} \\C\ar[rr]_{g} &&D&&C \ar[rr]_{g} & & D}\]
	We say that the left square is a \emph{Van Kampen square} if:
	\begin{enumerate}
		\item it is a pushout square;
		\item whenever the right cube has pullbacks as back and left faces, then its top face is a pushout if and only if the front and right faces are pullbacks.
	\end{enumerate}

	Pushout squares which enjoy the ``if'' half of this condition are called \emph{stable}.
\end{definition}

\begin{example}\label{ex:iso}
	In any category $\X$, a square as the one below on the left, in which $m$ is an isomorphism, is Van Kampen. Notice that, in this situation, $n$ must be an isomorphism too.
		\[\xymatrix@C=10pt@R=10pt{&&&&&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar[dl]_{m'} && B' \ar[dd]^{b} \ar[dl]_{n'} \\ A \ar[dd]_{m}\ar[rr]^{f}&&B\ar[dd]^{n}&&C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&&&&&A\ar[rr]|\hole^(.65){f} \ar[dl]^{m} && B \ar[dl]^{n} \\C\ar[rr]_{g} &&D&&C \ar[rr]_{g} & & D}\]

	Take a cube as the one above and suppose that its back faces are pullbacks. In particular we know that also $m'$ is an isomorphism.

	\smallskip \noindent 
	$(\Rightarrow)$ Suppose that the top face is a pushout, thus $n'$ is an isomorphism and the right square is a pullback. The thesis for the front face follows at once since $m', n', m$ and $m$ are isomorphisms and the back face a pullback.

	\smallskip \noindent 
	$(\Leftarrow)$ If all the vertical faces are pullbacks, then $m'$ and $n'$ are isomorphisms and this entails at once that the top face is a pushout.
\end{example}

@@

The following proposition establishes a key property of Van Kampen squares with a mono as a side: they are not only pushouts, but also pullbacks.
\begin{proposition}\label{prop:pbpo} Let $m\colon A\to C$ be a monomorphism in a category $\X$. Then every Van Kampen square
	\[\xymatrix{A\ar[r]^{g} \ar[d]_{m} & B \ar[d]^{n} \\ C \ar[r]_{f}  & D}\]
	is also a pullback square and $n$ is a monomorphism.
\end{proposition}
\begin{proof} Take the following cube:
	\[\xymatrix@C=13pt@R=13pt{&A\ar[dd]|\hole_(.65){\id{A}}\ar[rr]^{g} \ar[dl]_{\id{A}} && B \ar[dd]^{\id{B}} \ar[dl]_{\id{B}} \\ A  \ar[dd]_{m}\ar[rr]^(.65){g} & & B \ar[dd]_(.3){n}\\&A\ar[rr]|\hole^(.65){g} \ar[dl]^{m} && B \ar[dl]^{n} \\C \ar[rr]_{f} & & D}\]
	By construction the top face of the cube is a pushout and the back one a pullback. The left face is a pullback because $m$ is mono. Thus the Van Kampen property yields that the front and the right faces are pullbacks and the thesis follows. \qedhere 
\end{proof}

The previous proposition, in turn, allows us to establish the following results.
\begin{lemma}\label{lem:varie} Let $m:X\to Y$ be a monomorphism in a category $\X$ and suppose that the left square below is Van Kampen, while all the vertical faces in the right cube are pullbacks.
		\[\xymatrix@C=10pt@R=10pt{&&&&&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar[dl]_{m'} && B' \ar[dd]^{b} \ar[dl]_{n'} \\ A \ar[dd]_{m}\ar[rr]^{f}&&B\ar[dd]^{n}&&C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&&&&&A\ar[rr]|\hole^(.65){f} \ar[dl]^{m} && B \ar[dl]^{n} \\C\ar[rr]_{g} &&D&&C \ar[rr]_{g} & & D}\]
Supppose, moreover that $d:D'\to D$ is mono, then $d\leq n$ if and only if $c \leq m$.
\end{lemma}
\begin{remark}
	Recall that, given two monos $m:M\to X$ and $n:N\to X$ with the same codomain, $m\leq n$ means that there exists a, necessarily unique, $k:M\to N$ fitting in the triangle below:
	\[\xymatrix{M\ar@{.>}[rr]^{k}  \ar[dr]_{m}&& N \ar[dl]^{n}\\ & X}\]
	Notice that, if $m\leq n$ and $n\leq m$, then the arrow $k:M\to N$ is an isomorphism.  
\end{remark}
\begin{remark}
	Notice that, since $d$ is a mono and the front face a pullback, then $c$ is a monomorphism too.
\end{remark}

\begin{proof}
	$(\Rightarrow)$ By hypothesis there exists $k:D'\to B$ such that $n\circ k = d$. By \Cref{prop:pbpo}, the bottom face of the cube is a pullback. Thus there exists a unique $h:C'\to A$ as in the diagram below, implying the thesis.
		\[\xymatrix@C=40pt{C'  \ar@{.>}[d]^{h}\ar[r]^{g'} \ar@/_.5cm/[dd]_{c}& D' \ar@/^.5cm/[dd]^{d}\ar[d]_{k}\\ A \ar[d]^{m} \ar[r]^{f} & B \ar[d]_{n} \\  C \ar[r]_g & D}\]

	\smallskip \noindent 
	$(\Leftarrow)$ Let $h:C\to A$ be such that $c=m\circ h$. By the Van Kampen property the top face of the given cube is a pushout. Thus the dotted $k:D'\to B$ in the following diagram exists.
		\[\xymatrix@C=40pt{A' \ar@/_.5cm/[dd]_{a} \ar[d]^{m'} \ar[r]^{f'} & B' \ar[d]_{n'} \ar@/^.5cm/[dd]^{b} \\  C' \ar[d]^{h} \ar[r]_{g'} & D' \ar@{.>}[d]_{k}\\ A \ar[r]_f& B }\]
	The thesis now follows at once. \qedhere 
\end{proof}

Finally, we can show that stable pushouts enjoy a kind of \emph{pullback-pushout decomposition} property.

\begin{proposition}\label{prop:stab}Let $\X$ be a category  and suppose that, in the diagram below, the whole rectangle is a stable pushout and the right square a pullback.
	\[\xymatrix{X \ar[d]_{a} \ar[r]^{f}& \ar[r]^{g} Y \ar[d]^{b}& Z \ar[d]^{c}\\ A \ar[r]_{h}& B \ar[r]_{k}& C}\]
	If the arrow $k$ is a monomorphism,  then both squares are pushouts.
\end{proposition}

\begin{proof} We can begin noticinng that $g$, being the pullback of the monomorphism $k$, is monic too. Thus we can build the cube below, in which all the vertical faces are pullbacks.
	\[\xymatrix@C=20pt@R=10pt{ & &X\ar[dl]_{f} \ar[ddd]^(.333333){\id{X}}|(.666666)\hole\ar[rrr]^{a} &&&A \ar[ddd]^{\id{A}} \ar[dl]^{h}\\& Y\ar[dl]_{\id{Y}} \ar[ddd]|(.333333)\hole_{\id{Y}} &&&B \ar[ddd]^{\id{B}} \ar[dl]_{\id{B}} \\Y \ar[ddd]_{g} \ar[rrr]^{b}&&& B\ar[ddd]^{k}\\&&X \ar[rrr]|(.34)\hole^{a}|(.67)\hole \ar[dl]_{f}&&& A \ar[dl]^{h}\\ & Y  \ar[dl]_{g}&&& B \ar[dl]^{k}\\ Z\ar[rrr]_{c} &&& C}\]
	By hypothesis the face is a stable pushout and so its top one is a pushout. \Cref{lem:po1} now entails that the right half of the rectangle with which we have started is a pushout too. \qedhere 
\end{proof}

@@

In this section we will define the notion of $\mathcal{M}, \mathcal{N}$-adhesivity and explore some of the consequence of such a property. Let us start fixing some terminology.

\begin{definition}\
	Let $\mathcal{M}$  be a class of arrows of a category $\X$. We say that  $\mathcal{M}$ is
	\begin{itemize}
		\item 
		\emph{stable under pushouts (pullbacks)} if for every pushout (pullbacks) square 
		\[\xymatrix{A\ar[r]^f  \ar[d]_{m}& B \ar[d]^n \\ C \ar[r]_g & D}\]
		if $m \in \mathcal{M}$ ($n\in \mathcal{M}$) then $n \in \mathcal{M}$ ($m \in \mathcal{M}$);
		\item \emph{closed under composition} if $g, f\in \mathcal{M}$ implies $g\circ f\in \mathcal{M}$ whenever $g$ and $f$ are composable;
		\item \emph{closed under decomposition} if $g\circ f\in \mathcal{M}$ and $g\in \mathcal{M}$ implies $f\in \mathcal{M}$.
	\end{itemize}
\end{definition}

\begin{remark}Clearly, ``decomposition'' corresponds to ``left cancellation'', but we prefer to stick to the name commonly used in literature.
\end{remark}

We are now ready to give the definition of $\mathcal{M}$-adhesive category	
\begin{definition}[\cite{azzi2019essence,heindel2009category}]
	Let $\X$ be a category and consider a subclass $\mathcal{M}$ of the class $\mathcal{M}(\X)$ of monomorphisms such that:
	\begin{enumerate}
		\item $\mathcal{M}$ contains all isomorphisms and is closed under composition;
		\item $\mathcal{M}$ is stable under pullbacks and pushouts.
	\end{enumerate}
	We say that $\X$ is \emph{$\mathcal{M}$-adhesive} if
	\begin{enumerate}
		\item for every $m\colon X\to Y$ in $\mathcal{M}$ and $g\colon Z\to Y$, a pullback square
		\[\xymatrix{P\ar[r]^p \ar[d]_{n}& X \ar[d]^{m}\\ Z \ar[r]_g& Y}\]
		exists, such pullbacks will be called \emph{$\mathcal{M}$-pullbacks};
		\item for every $m\colon X\to Y$ in $\mathcal{M}$ and $n\colon X\to Z$, a pushout square
		\[\xymatrix{X \ar[r]^n \ar[d]_{m}& Z \ar[d]^{q}\\ Y\ar[r]_p &Q}\]
		exists, such pushouts  will be called \emph{$\mathcal{M}$-pushouts}; 
		\item  $\mathcal{M}$-pushouts are Van Kampen squares.
	\end{enumerate}
\end{definition}

\begin{remark}\label{rem:diff}Our notion of $\mathcal{M}$-adhesivity is slightly different from the one of \cite{azzi2019essence}: in that paper, $\mathcal{M}$ is assumed to be only stable under pullbacks. Notice, however, that if $\mathcal{M}$ contains all split monos, then stability under pushouts can be deduced from the other axioms \cite[Prop.~$5.1.21$]{castelnovo2023thesis}.
\end{remark}

\begin{remark}\label{rem:salva} 
	\emph{Adhesivity} and \emph{quasiadhesivity} as defined in \cite{lack2005adhesive,garner2012axioms}  coincide with  $\mathcal{A}(\X) $-adhesivity and $\mathcal{R}(\X)$-adhesivity, respectively. 
\end{remark}

@@

\begin{proposition}\label{prop:deco}Let  $\mathcal{M}$ be a class of monomorphisms containing all isomorphisms and stable under pullbacks. For every arrow $f: X\to Y$ and monomorphism $m: Y\to Z$, if $m\circ f \in\mathcal{M}$ then $f\in \mathcal{M}$.
\end{proposition}
\begin{proof}Take the diagram
	\[\xymatrix{X \ar[d]_{\id{X}}\ar[r]^{f}& Y \ar[r]^{\id{Y}}  \ar[d]_{\id{Y}}& Y \ar[d]^{m}\\
		X \ar[r]_{f}& Y \ar[r]_{m} & Z}\]
	Since $m$ is mono the right square is a pullback, the thesis now follows from \Cref{lem:pb1}.
\end{proof}
\begin{corollary}\label{cor:deco}
	In every $\mathcal{M}$-adhesive category $\X$, the class $\mathcal{M}$ is closed under decomposition.
\end{corollary}

@@

From \Cref{prop:pbpoad}, in turn, we can derive the following corollaries.
\begin{corollary}\label{cor:rego}
	In a $\mathcal{M}$-adhesive category $\X$, every $m\in\mathcal{M}$ is a regular mono.
\end{corollary}
\begin{proof}
Let $m$ be an element of $\mathcal{M}$ and consider its pushout along itself.
\[\xymatrix{X\ar[r]^m \ar[d]_m& Y\ar[d]^{f}\\Y \ar[r]_g & Z}\]
By \Cref{prop:pbpoad} this square is a pullback, proving that $m$ is the equalizer of the arrows $f,g\colon Y\rightrightarrows Z$. \qedhere 
\end{proof}

@@

$\mathcal{M}$-adhesivity is well-behaved with respect to  the comma construction \cite{mac2013categories}, as shown by the following theorem.
\begin{theorem}[\cite{CastelnovoGM22,ehrig2006fundamentals,lack2005adhesive}]\label{lem:comma}
	Let $\A$ and $\B$ be respectively an $\mathcal{M}$-adhesive and an $\mathcal{M}'$-adhesive category. Let also $L:\A\rightarrow \C$ be a functor that preserves $\mathcal{M}$-pushouts, and  $R:\B\rightarrow \C$ be a functor which preserves pullbacks.Then $\comma{L}{R}$ is $\cma{M}{M'}$-adhesive, where 
	\[
	\cma{M}{M}':=\{(h,k)\in \mathcal{A}(\comma{L}{R}) \mid h\in \mathcal{M}, k\in \mathcal{M}'\}\]
\end{theorem}

@@

\begin{theorem}[\cite{CastelnovoGM22,ehrig2006fundamentals,lack2005adhesive}]\label{thm:functors}
If $\X$ is an $\mathcal{M}$-adhesive category, then for every small category $\Y$, the category $\X^\Y$  of functors $\Y\to \X$ is $\mathcal{M}^{\Y}$-adhesive, where
\[\mathcal{M}^{\Y}:=\{\eta \in \mathcal{A}(\X^\Y) \mid \eta_Y \in \mathcal{M} \text{ for every } Y\in \Y\}\]
\end{theorem}

@@

@@

A \emph{left-linear DPO-rewriting system} is a pair $(\X, \R)$ where $\X$ is a $\mathcal{M}$-adhesive category and $R$ is a set of left $\mathcal{M}$-linear rules. $(\X, \R)$ will be called \emph{linear} if every rule in $R$ is so.

>>>>

\usepackage[all]{xy}
\newdir^{ (}{{}*!/-5pt/@^{(}}
\SelectTips{cm}{}
% derivations and labels
\usepackage{proof}
\newcommand{\lab}[1]{\ensuremath{\mathsf{{#1}}}}
\newcommand{\slab}[1]{\ensuremath{\scriptstyle{\mathsf{{#1}}}}}

%frecce
\newcommand{\mor}{\mathsf{Mor}}
\newcommand{\mon}{\mathsf{Mono}}
\newcommand{\reg}{\mathsf{Reg}}

@@

%identità
\newcommand{\id}[1]{\mathsf{id}_{#1}}
%codominio
\newcommand{\cod}[1]{\mathsf{cod}({#1})}
	%Variabili categorie
\def\A{\textbf {\textup{A}}}
%mono
\newcommand{\mto}[0]{\scalebox{1}{$\rightarrowtail$}}

@@

\title{Switch equivalence and weak prime domains for fusions}
\author{Paolo Baldan\and Davide Castelnovo\and  Andrea Corradini\and Fabio Gadducci}

@@

This first section is devoted to recall the definition and the basic theory of \emph{$\mathcal{M}$-adhesive categories} \cite{azzi2019essence,ehrig2012,ehrig2014adhesive,lack2005adhesive}.

\begin{notation} 
We stipulate here some notational conventions which will be used throughout this paper. 
\begin{itemize}\item 
	Given a category $\X$ we will not distinguish notationally between $\X$ and its class of objects: so that ``$X\in \X$'' means that $X$ belongs to the class of objects of $\X$.  
	\item 
	If $1$ is a terminal object in a category $\X$,  the unique arrow $X\to 1$ from another object $X$ will be denoted by $!_X$. Similarly, if $0$ is initial in $\X$ then $?_X$ will denote the unique arrow $0\to X$. %When $\X$ is $\Set$ and $1$ is a singleton, $\delta_x$ will denote the arrow $1\to X$ with value $x\in X$.
	\item  $\mor(\X)$, $\mon(\X)$ and $\reg(\X)$ will denote the class of all arrows, monos and regular monos of $\X$, respectively.

	\item Given an integer $n\in \mathbb{Z}$, $[0,n]$ will denote the set
	\[[0,n]:=\{x\in \mathbb{N}\mid x\leq n\}\]
	Indeed, if $n<0$, then $[0,n]=\emptyset$.
\end{itemize}
\end{notation}

\subsection{The Van Kampen property}
The key property that $\mathcal{M}$-adhesive categories enjoy is given by  the so-called \emph{Van Kampen condition} \cite{brown1997van,johnstone2007quasitoposes,lack2005adhesive}. We will recall it and examine some of its consequences. First of all we need to recall some terminology and facts regarding subclasses of $\mor(\X)$.

\begin{definition}
	Let $\X$ be a category and $\mathcal{A}$ a  subclass of $\mor(\X)$. We say that  $\mathcal{A}$ is
	\begin{itemize}
		\item 
		\emph{stable under pushouts (pullbacks)} if for every pushout (pullbacks) square 
		\[\xymatrix{A\ar[r]^f  \ar[d]_{m}& B \ar[d]^n \\ C \ar[r]_g & D}\]
		if $m \in \mathcal{A}$ ($n\in \mathcal{A}$) then $n \in \mathcal{A}$ ($m \in \mathcal{A}$);
		\item \emph{closed under composition} if $g, f\in \mathcal{A}$ implies $g\circ f\in \mathcal{A}$ whenever $g$ and $f$ are composable;
		\item \emph{closed under decomposition} if $g\circ f, g\in \mathcal{A}$ implies $f\in \mathcal{A}$.
	\end{itemize}
\end{definition}

\begin{remark}In the previous definition, ``decomposition'' corresponds to ``left cancellation'', but we prefer to stick to the name commonly used in literature (see e.g.~\cite{habel2012mathcal}).
\end{remark}

So equipped, we can introduce the notion of \emph{$\mathcal{A}$-Van Kampen square}.
\begin{definition}[Van Kampen property] Let $\X$ be a category  and consider the two diagrams below
	\[\xymatrix@C=10pt@R=10pt{&&&&&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar[dl]_{m'} && B' \ar[dd]^{b} \ar[dl]_{n'} \\ A \ar[dd]_{m}\ar[rr]^{f}&&B\ar[dd]^{n}&&C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&&&&&A\ar[rr]|\hole^(.65){f} \ar[dl]^{m} && B \ar[dl]^{n} \\C\ar[rr]_{g} &&D&&C \ar[rr]_{g} & & D}\]

	Given  a class of arrows $\mathcal{A}\subseteq \mor(\X)$, we say that the left square \emph{has the Van Kampen property relatively to $\mathcal{A}$}, or that it  is a \emph{$\mathcal{A}$-Van Kampen square} if:
	\begin{enumerate}
		\item is a pushout square;
		\item 	whenever the right cube has pullbacks as back and left faces and the vertical arrows belong to $\mathcal{A}$, then its top face is a pushout if and only if the front and right faces are pullbacks.
	\end{enumerate}
	Pushout squares which enjoy the ``if'' half of the second point of the condition above are called \emph{$\mathcal{A}$-stable}.

We will call a $\mor(\X)$-Van Kampen ($\mor(\X)$-stable) square simply a \emph{Van Kampen} (\emph{stable}) square.
\end{definition}

@@

The following proposition establishes a key property of $\mathcal{A}$-Van Kampen squares with a mono as a side: they are not only pushouts, but also pullbacks.
\begin{proposition}\label{prop:pbpo} Let $\mathcal{A}$ be a class of arrows stable under pushouts and containing all the isomorphisms.  If $m\colon A\to C$ is mono and belongs to $\mathcal{A}$, then every $\mathcal{A}$-Van Kampen square
	\[\xymatrix{A\ar[r]^{g} \ar[d]_{m} & B \ar[d]^{n} \\ C \ar[r]_{f}  & D}\]
	is also a pullback square and $n$ is a monomorphism.
\end{proposition}
\begin{proof} We can start considering the cube below and noticing that $n$, being the pushout of $m\in \mathcal{M}$, belongs to $\mathcal{A}$ too.
	\[\xymatrix@C=13pt@R=13pt{&A\ar[dd]|\hole_(.65){\id{A}}\ar[rr]^{g} \ar[dl]_{\id{A}} && B \ar[dd]^{\id{B}} \ar[dl]_{\id{B}} \\ A  \ar[dd]_{m}\ar[rr]^(.65){g} & & B \ar[dd]_(.3){n}\\&A\ar[rr]|\hole^(.65){g} \ar[dl]^{m} && B \ar[dl]^{n} \\C \ar[rr]_{f} & & D}\]
	By construction the top face of the cube is a pushout and the back one a pullback. The left face is a pullback because $m$ is mono. Thus the $\mathcal{A}$-Van Kampen property yields that the front and the right faces are pullbacks. 
\end{proof}

The previous proposition, in turn, allows us to establish the following results.
\begin{lemma}\label{lem:varie}Let $\mathcal{A}$ be a class of arrows stable under pullbacks, pushouts and containing all isomorphisms.  Suppose that, the left square below is $\mathcal{A}$-Van Kampen, while the vertical faces in the right cube are pullbacks.
		\[\xymatrix@C=10pt@R=10pt{&&&&&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{g'} \ar[dl]_{m'} && B' \ar[dd]^{b} \ar[dl]_{n'} \\ A \ar[dd]_{m}\ar[rr]^{g}&&B\ar[dd]^{n}&&C'  \ar[dd]_{c}\ar[rr]^(.7){f'} & & D' \ar[dd]_(.3){d}\\&&&&&A\ar[rr]|\hole^(.65){g} \ar[dl]^{m} && B \ar[dl]^{n} \\C\ar[rr]_{f} &&D&&C \ar[rr]_{f} & & D}\]
Supppose, moreover that $m:A\to C$ and $d:D'\to D$ are mono and that $d$ belongs to $\mathcal{A}$. Then $d\leq n$ if and only if $c \leq m$.
\end{lemma}

\begin{remark}
	Recall that, given two monos $m:M\to X$ and $n:N\to X$ with the same codomain, $m\leq n$ means that there exists a, necessarily unique and necessarily mono, $k:M\to N$ fitting in the triangle below:
	\[\xymatrix{M\ar[rr]^{k}  \ar[dr]_{m}&& N \ar[dl]^{n}\\ & X}\]
	Notice that, if $m\leq n$ and $n\leq m$, then the arrow $k:M\to N$ is an isomorphism.  
\end{remark}
\begin{remark}
	Notice that, since $d$ is a mono and the vertical faces are pullbacks, then $a, b$ and $c$ are monomorphisms too. Moreover, $n$ is mono by \Cref{prop:pbpo}, so that even $m'$ and $n'$ are monos.
\end{remark}

\begin{proof}
	$(\Rightarrow)$ By hypothesis there exists $k:D'\to B$ such that $n\circ k = d$. By \Cref{prop:pbpo}, the bottom face of the cube is a pullback. Thus there exists a unique $h:C'\to A$ as in the diagram below, implying the thesis.
		\[\xymatrix@C=40pt{C'  \ar[d]^{h}\ar[r]^{f'} \ar@/_.5cm/[dd]_{c}& D' \ar@/^.5cm/[dd]^{d}\ar[d]_{k}\\ A \ar[d]^{m} \ar[r]^{g} & B \ar[d]_{n} \\  C \ar[r]_f & D}\]

	\smallskip \noindent 
	$(\Leftarrow)$ Let $h:C\to A$ be such that $c=m\circ h$. By the $\mathcal{A}$-Van Kampen property the top face of the given cube is a pushout. Thus the dotted $k:D'\to B$ in the following diagram exists.
		\[\xymatrix@C=40pt{A' \ar@/_.5cm/[dd]_{a} \ar[d]^{m'} \ar[r]^{g'} & B' \ar[d]_{n'} \ar@/^.5cm/[dd]^{b} \\  C' \ar[d]^{h} \ar[r]_{f'} & D' \ar@{.>}[d]_{k}\\ A \ar[r]_g& B }\]
Moreover, by construction we have
\[\begin{split}
n\circ k \circ n' &= n\circ b\\&=d\circ n'\\&\\&
\end{split} \qquad \begin{split}
n\circ k \circ f' &= n\circ g\circ h\\&=f\circ m\circ h\\&=f\circ c\\&= d\circ f' \end{split}\]
We can therefore conclude that $n\circ k =d$. \qedhere 
\end{proof}

Finally, we can show that $\mathcal{A}$-stable pushouts enjoy a kind of \emph{pullback-pushout decomposition} property.

\begin{proposition}\label{prop:stab}Let $\X$ be a category and $\mathcal{A}$ a class of arrows stable under pullbacks. Suppose that, in the diagram below, the whole rectangle is an $\mathcal{A}$-stable pushout and the right square a pullback.
	\[\xymatrix{X \ar[d]_{a} \ar[r]^{f}& \ar[r]^{g} Y \ar[d]^{b}& Z \ar[d]^{c}\\ A \ar[r]_{h}& B \ar[r]_{k}& C}\]
	If the arrow $k$ is in $\mathcal{A}$ and it is a monomorphism,  then both squares are pushouts.
\end{proposition}

\begin{proof} We can begin noticing that $g$, being the pullback of $k$, is mono and in $\mathcal{A}$ too. Thus we can build the cube below, in which all the vertical faces are pullbacks, entailing that all the vertical arrows are in $\mathcal{A}$.
	\[\xymatrix@C=20pt@R=10pt{ & &X\ar[dl]_{f} \ar[ddd]^(.333333){\id{X}}|(.666666)\hole\ar[rrr]^{a} &&&A \ar[ddd]^{\id{A}} \ar[dl]^{h}\\& Y\ar[dl]_{\id{Y}} \ar[ddd]|(.333333)\hole_{\id{Y}} &&&B \ar[ddd]^{\id{B}} \ar[dl]_{\id{B}} \\Y \ar[ddd]_{g} \ar[rrr]^{b}&&& B\ar[ddd]^{k}\\&&X \ar[rrr]|(.34)\hole^{a}|(.67)\hole \ar[dl]_{f}&&& A \ar[dl]^{h}\\ & Y  \ar[dl]_{g}&&& B \ar[dl]^{k}\\ Z\ar[rrr]_{c} &&& C}\]
	By hypothesis the face is an $\mathcal{A}$-stable pushout and so its top one is a pushout. Using \Cref{lem:po1} we can conclude that the right half of the rectangle with which we have started is a pushout too. \qedhere 
\end{proof}

@@

In this section we will define the notion of $\mathcal{M}$-adhesivity \cite{azzi2019essence,ehrig2012,ehrig2014adhesive,heindel2009category,lack2005adhesive} and explore some of the consequence of such a property.

\begin{definition}[$\mathcal{M}$-adhesive category]
	Let $\X$ be a category and consider a subclass $\mathcal{M}$ of the class $\mon(\X)$ of monomorphisms such that:
	\begin{enumerate}
		\item $\mathcal{M}$ contains all isomorphisms and is closed under composition;
		\item $\mathcal{M}$ is stable under pullbacks and pushouts.
	\end{enumerate} 
	We will use $m\colon X\mto Y$ to denote that an arrow $m:X\to Y$ belongs to $\mathcal{M}$. $\X$ is said to be \emph{$\mathcal{M}$-adhesive} if
	\begin{enumerate}
		\item for every $m\colon X\mto Y$ in $\mathcal{M}$ and $g\colon Z\to Y$, a pullback square
		\[\xymatrix{P\ar[r]^p \ar@{>->}[d]_{n}& X \ar@{>->}[d]^{m}\\ Z \ar[r]_g& Y}\]
		exists, such pullbacks will be called \emph{$\mathcal{M}$-pullbacks};
		\item for every $m\colon X\mto Y$ in $\mathcal{M}$ and $f\colon X\to Z$, a pushout square
		\[\xymatrix{X \ar[r]^f \ar@{>->}[d]_{m}& Z \ar[d]^{q}\\ Y\ar[r]_p &Q}\]
		exists, such pushouts  will be called \emph{$\mathcal{M}$-pushouts}; 
		\item  $\mathcal{M}$-pushouts are $\mathcal{M}$-Van Kampen squares.
	\end{enumerate}

A category $\X$ is said to be \emph{strictly $\mathcal{M}$-adhesive} if $\mathcal{M}$-pushouts are Van Kampen squares.	
\end{definition}

\begin{remark}\label{rem:diff}Our notion of $\mathcal{M}$-adhesivity follows \cite{ehrig2012,ehrig2014adhesive} and is different from the one of \cite{azzi2019essence}. What is called $\mathcal{M}$-adhesivity in that paper corresponds to our strict $\mathcal{M}$-adhesivity. Moreover, in \cite{azzi2019essence} the class $\mathcal{M}$ is assumed to be only stable under pullbacks. However, if $\mathcal{M}$ contains all split monos, then stability under pushouts can be deduced from the other axioms \cite[Prop.~$5.1.21$]{castelnovo2023thesis}.
\end{remark}

\begin{remark}\label{rem:salva} 
	\emph{Adhesivity} and \emph{quasiadhesivity} as defined in \cite{lack2005adhesive,garner2012axioms}  coincide with  strict $\mon(\X) $-adhesivity and strict $\reg(\X)$-adhesivity, respectively. 
\end{remark}

@@

\begin{proposition}\label{prop:deco}Let  $\mathcal{A}$ be a class of arrows stable under pullbacks. For every arrow $f\colon X\to Y$ and monomorphism $m\colon Y\to Z$, if $m\circ f \in\mathcal{A}$ then $f\in \mathcal{A}$.
\end{proposition}
\begin{proof}Take the diagram
	\[\xymatrix{X \ar[d]_{\id{X}}\ar[r]^{f}& Y \ar[r]^{\id{Y}}  \ar[d]_{\id{Y}}& Y \ar[d]^{m}\\
		X \ar[r]_{f}& Y \ar[r]_{m} & Z}\]
	Since $m$ is mono the right square is a pullback, while the left square is a pullback by construction. By \Cref{lem:pb1} the whole rectangle is a pullback and the thesis follows.
\end{proof}
\begin{corollary}\label{cor:deco}
	In every $\mathcal{M}$-adhesive category $\X$, the class $\mathcal{M}$ is closed under decomposition.
\end{corollary}

@@

From \Cref{prop:pbpoad}, in turn, we can derive the following corollaries.
\begin{corollary}\label{cor:rego}
	In a $\mathcal{M}$-adhesive category $\X$, every $m\in\mathcal{M}$ is a regular mono.
\end{corollary}
\begin{proof}
Let $m$ be an element of $\mathcal{M}$ and consider its pushout along itself.
\[\xymatrix{X\ar@{>->}[r]^m \ar@{>->}[d]_m& Y\ar@{>->}[d]^{f}\\Y \ar@{>->}[r]_g & Z}\]
By \Cref{prop:pbpoad} this square is a pullback, proving that $m$ is the equalizer of the arrows $f,g\colon Y\rightrightarrows Z$. \qedhere 
\end{proof}

@@

$\mathcal{M}$-adhesivity is well-behaved with respect to  the comma construction \cite{mac2013categories}, as shown by the following theorem.
\begin{theorem}[\cite{ehrig2006fundamentals,lack2005adhesive}]\label{lem:comma}
	Let $\A$ and $\B$ be respectively an $\mathcal{M}$-adhesive and an $\mathcal{M}'$-adhesive category. Let also $L:\A\rightarrow \C$ be a functor that preserves $\mathcal{M}$-pushouts, and  $R:\B\rightarrow \C$ be a functor which preserves pullbacks.Then $\comma{L}{R}$ is $\cma{M}{M'}$-adhesive, where 
	\[
	\cma{M}{M}':=\{(h,k)\in \mathcal{A}(\comma{L}{R}) \mid h\in \mathcal{M}, k\in \mathcal{M}'\}\]
\end{theorem}

@@

\begin{theorem}[\cite{ehrig2006fundamentals,lack2005adhesive}]\label{thm:functors}
If $\X$ is an $\mathcal{M}$-adhesive category, then for every small category $\Y$, the category $\X^\Y$  of functors $\Y\to \X$ is $\mathcal{M}^{\Y}$-adhesive, where
\[\mathcal{M}^{\Y}:=\{\eta \in \mathcal{A}(\X^\Y) \mid \eta_Y \in \mathcal{M} \text{ for every } Y\in \Y\}\]
\end{theorem}

@@

\begin{example}
	\todo{grafi semplici}
\end{example}

@@

A \emph{left-linear DPO-rewriting system} is a pair $(\X, \R)$ where $\X$ is a $\mathcal{M}$-adhesive category and $R$ is a set of left $\mathcal{M}$-linear rules. $(\X, \R)$ will be called \emph{linear} if every rule in $R$ is $\mathcal{M}$-linear.
